Efficient UQ for chemical systems using active \author{Manav Vohra, Alen Alexanderian, Hayley Guy, Sankaran Mahadevan} subspaces

### 2018-08-06 Paper/app.tex
..

The 1-dimensional surrogate ($\tilde{G}$) is investigated for its accuracy as well as the ability to capture the 
uncertainty in the
model output in two ways. Firstly, we estimate the relative L-2 norm of the discrepancy~($\varepsilon_d$)
between estimates of ignition delay in the case of H$_2$/O$_2$ reaction, obtained using the 
model output and the surrogate in the following equation:
%
\be
\varepsilon_d = \frac{\|G(\bm{\xi}) - \tilde{G}(\bm{\xi})\|_2}{\|G(\bm{\xi})\|_2}
\ee
%
The relative norm, $\varepsilon_d$ was estimated to be 8.50$\times10^{-2}$ based on a validation test data
comprising 1000 model evaluations. Hence, in the norm-sense, it can be said that the 1-dimensional surrogate is 
reasonably accurate.

Secondly, we verify the accuracy of the surrogate in a probabilistic setting. In particular, we compare a histogram
plot of the model evaluations for ignition delay in the validation set with a PDF of the same using the 1-dimensional
surrogate as shown in Figure~\ref{fig:pdf_33D}.
%
\begin{figure}[htbp]
 \begin{center}
  \includegraphics[width=0.45\textwidth]{./Figures/pdf_comp_id_1e4}
\caption{A comparison of the histogram plot based on model evaluations and the PDF based on surrogate
predictions for ignition delay. The two plots are based on samples in the validation set.}
\label{fig:pdf_33D}
\end{center}
\end{figure}
%
In the above comparison, it is observed that the peak of the PDF generated using the 1-dimensional surrogate in the
active subspace is consistent with the modal estimate based on the histogram. Moreover, the spread in the ignition
 delay is reasonably captured by the PDF. Hence, the surrogate is also verified in a probabilistic sense.

>>>>

The 1-dimensional surrogate ($\tilde{G}$) is investigated for its accuracy as well as the ability to capture the 
uncertainty in the
model output in two ways. Firstly, we estimate the relative L-2 norm of the discrepancy~($\varepsilon_d$)
between estimates of ignition delay in the case of H$_2$/O$_2$ reaction, obtained using the 
model output and the surrogate in the following equation:
%
\be
\varepsilon_d = \frac{\|G(\bm{\xi}) - \tilde{G}(\bm{\xi})\|_2}{\|G(\bm{\xi})\|_2}
\ee
%
The relative norm, $\varepsilon_d$ was estimated to be 8.50$\times10^{-2}$ based on a cross-validation test data
comprising 10$^{4}$ model evaluations. Hence, in the norm-sense, it can be said that the 1-dimensional surrogate is 
reasonably accurate.

Secondly, we verify the accuracy of the surrogate in a probabilistic setting. In particular, we compare a histogram
plot of the model evaluations (in the cross-validation set) with a PDF evaluated on the same set of samples
using the surrogate, $\tilde{G}$ in Figure~\ref{fig:pdf_33D}. The PDF is evaluated using surrogates constructed for
a 1-dimensional as well as a 2-dimensional active subspace. 
%
\begin{figure}[htbp]
 \begin{center}
  \includegraphics[width=0.45\textwidth]{./Figures/pdf_comp_id_1e4}
\caption{A comparison of the histogram plot based on model evaluations and the PDF based on surrogate
predictions for ignition delay. The two plots are based on samples in the validation set.}
\label{fig:pdf_33D}
\end{center}
\end{figure}
%
In the above comparison, it is observed that the two PDFs are almost identical, thereby confirming a 1-dimensional
active subspace. The modal estimate from either PDF is found to be in agreement with that
based on the histogram plot. Moreover, the spread in the ignition
delay is reasonably captured by the PDFs.

Based on these observations, it can be said that a 1-dimensional surrogate can be used to reasonably approximate the
uncertainty in the ignition delay due to uncertain rate-controlling parameters used in the H$_2$/O$_2$ reaction
mechanism.

### 2018-08-06 Paper/assa.tex
..

\subsection{Active subspaces}
\label{sub:ac}
As mentioned in the introduction, the active subspace is low-dimensional subspace
that constitutes important directions in a model's input
space~\cite{Constantine:2015}. The effective variability in the model output, $f$
due to uncertain inputs, $\bm{\theta}$ in the physical space, is predominantly captured
along these directions. Each component of $\bm{\theta}$
is mapped to a corresponding value in $\Omega\in\mathbb{R}^{N_p}$
($N_p$: number of uncertain inputs)
using a linear transformation.  The set of canonical random variables, corresponding
to $\theta_i$'s, in $\Omega$ is referred to as $\bm{\xi}$.
Hence, from this point onwards, we denote $f$ as $f(\bm{\xi})$.
Mathematically, for a given $f$, and a set of inputs, $\bm{\xi}$,
these directions are the eigenvectors of a positive
semi-definite matrix, $\mat{C}$, defined as follows: 
%
\be
\mat{C} = \int_\Omega (\nabla_{\bm{\xi}}f)(\nabla_{\bm{\xi}}f)^\top dP_{\bm{\xi}}
\label{eq:C}
\ee
%
where 
$dP_{\bm{\xi}}$ = $p(\bm{\xi})d\bm{\xi}$, where $p(\bm{\xi})$ is the joint probability
distribution of $\bm{\xi}$. Note that the function, $f$ is considered to be continuous 
and differentiable in the physical space, $\bm{\theta}$. %Hence, it is possible that a given $f(\bm{\xi})$ might not admit an active
%subspace. However, it is of remarkable interest to investigate if one exists
%since the subspace could be exploited to reduce the dimensionality of the
%problem and hence the associated computational effort.
Here 
$\nabla_{\bm{\xi}}f$ denotes the gradient vector. 
%with individual components
%being partial derivatives of $f$ with respect to the $i^\text{th}$ input, $\xi_i$. 
Since $\mat{C}$ is symmetric and
positive semi-definite, it admits a spectral decomposition:
%
\be
\mat{C} = \mat{W}\mat{\Lambda}\mat{W}^\top
\ee
%
where $\mat{\Lambda}$ = diag($\lambda_1,\ldots,\lambda_{N_p}$) with
$\lambda_i$'s sorted in descending order
\[
     \lambda_1 \geq \lambda_2 \geq \cdots \geq \lambda_\Np \geq 0.
\] 
The eigenpairs are partitioned about the $i^{\text{th}}$ eigenvalue such that
the ratio, $\left(\frac{\lambda_i}{\lambda_{i+1}}\right)$ $\gg$ 1 as follows:
\be
 \mat{W} = [\mat{W}_1~\mat{W}_2],~~\mat{\Lambda} = \begin{bmatrix}\mat{\Lambda}_1 & \\  &
  \mat{\Lambda}_2 \\
\end{bmatrix}
\ee
 %
 where $\mat{W}_1$ comprises the dominant eigenspace of $\mat{C}$ regarded as
the active subspace, and $\mat{\Lambda}_1$ is the corresponding set of
eigenvalues. Once the active subspace is computed, dimension reduction is
accomplished by transforming the set of canonical random variables, $\bm{\xi}$
into a reduced set, $\mat{W}_1^\top \bm{\xi}$, referred to as the set of active
variables. The number of active variables is thus equal to the number
of dominant eigenvectors. The model output in the original space, $f(\bm{\xi}(\theta))$,
is expressed in terms of the active variables as $G(\mat{W}_1^\top \bm{\xi})$.
In practice, the matrix, $\mat{C}$ is typically
approximated using pseudo-random sampling techniques such as Monte Carlo and
Latin hypercube sampling. The integral in~\eqref{eq:C} is replaced with a
summation as follows:
 %
 \be
 \mat{C}\approx \hat{\mat{C}} = \frac{1}{N}\sum\limits_{i=1}^{N} 
 (\nabla_{\bm{\xi}}f(\bm{\xi}_i))(\nabla_{\bm{\xi}}f(\bm{\xi}_i))^\top
 = \hat{\mat{W}}\hat{\mat{\Lambda}}\hat{\mat{W}}^\top
\label{eq:chat}
 \ee
 %
Clearly, the amount of computational effort associated with constructing the matrix,
$\hat{\mat{C}}$ scales with the number of samples, $N$. Hence, an iterative
computational approach is adopted in this work to ensure that $N$ is sufficiently
large so that the dominant eigenvectors in $\hat{\mat{W}}_1$ are approximated
with reasonable accuracy as discussed later in Section~\ref{sec:method}.

@@

Combining Proposition~\ref{prp:dgsm_bound} and~\eqref{equ:sobol_bound}, shows
an interesting link between the activity scores and total Sobol' indices.
Specifically, by computing the activity scores, we can indetify unimportant
parameters.  
One can attempt to reduce parameter dimension by fixing
unimportant parameters at nominal values. Suppose $\xi_i$ is
deemed unimportant as a result of a global sensitivity analysis
based on computing the activity scores. We want to understand
what is the approximation error that occurs once we fix $\xi_i$ 
at a nominal value.
%Let $\I \subset \{1, \ldots,
%Np\}$ be an index set that indexes the unimportant parameters in the parameter
%vector $\vec{\xi}$.  Letting $|\I|$ denote the number of elements of $\I$, 
%we denote by $\vec{z} \in \R^{|\I|}$ a generic vector of nominal values for
%the unimportant parameters $\vec\xi_\I$, and let $\vec{y}$ be a generic vector of
%important variables, $\vec\xi_{\I^c}$,
%where $\I^c$ denotes the complement of $\I$ in $\{1, \ldots, \Np\}$. 
%For a vector $\xi$, we let $\xin$ denote the vector obtained by removing
%$\xi_i$.

>>>>

\subsection{Active subspaces}
\label{sub:ac}
As mentioned in the introduction, the active subspace is low-dimensional subspace
that constitutes important directions in a model's input
space~\cite{Constantine:2015}. The effective variability in the model output, $f$
due to uncertain inputs, $\bm{\theta}$ in the physical space, is predominantly captured
along these directions. Each component of $\bm{\theta}$
is mapped to a corresponding value in $\Omega\in\mathbb{R}^{N_p}$
($N_p$: number of uncertain inputs)
using a linear transformation.  The set of canonical random variables, corresponding
to $\theta_i$'s, in $\Omega$ is referred to as $\bm{\xi}$.
Hence, from this point onwards, we denote $f$ as $f(\bm{\xi})$.
Mathematically, for a given $f$, and a set of inputs, $\bm{\xi}$,
these directions are the eigenvectors of a positive
semi-definite matrix, $\mat{C}$, defined as follows: 
%
\be
\mat{C} = \int_\Omega (\nabla_{\bm{\xi}}f)(\nabla_{\bm{\xi}}f)^\top dP_{\bm{\xi}}
\label{eq:C}
\ee
%
where 
$dP_{\bm{\xi}}$ = $p(\bm{\xi})d\bm{\xi}$, where $p(\bm{\xi})$ is the joint probability
distribution of $\bm{\xi}$. Note that the function, $f$ is considered to be continuous 
and differentiable in the physical space, $\bm{\theta}$. 
%Hence, it is possible that a given $f(\bm{\xi})$ might not admit an active
%subspace. However, it is of remarkable interest to investigate if one exists
%since the subspace could be exploited to reduce the dimensionality of the
%problem and hence the associated computational effort.
Here 
$\nabla_{\bm{\xi}}f$ denotes the gradient vector. 
%with individual components
%being partial derivatives of $f$ with respect to the $i^\text{th}$ input, $\xi_i$. 
Since $\mat{C}$ is symmetric and
positive semi-definite, it admits a spectral decomposition:
%
\be
\mat{C} = \mat{W}\mat{\Lambda}\mat{W}^\top
\ee
%
where $\mat{\Lambda}$ = diag($\lambda_1,\ldots,\lambda_{N_p}$) with
$\lambda_i$'s sorted in descending order
\[
     \lambda_1 \geq \lambda_2 \geq \cdots \geq \lambda_\Np \geq 0.
\] 
The eigenpairs are partitioned about the $i^{\text{th}}$ eigenvalue such that
the ratio, $\left(\frac{\lambda_i}{\lambda_{i+1}}\right)$ $\gg$ 1 as follows:
\be
 \mat{W} = [\mat{W}_1~\mat{W}_2],~~\mat{\Lambda} = \begin{bmatrix}\mat{\Lambda}_1 & \\  &
  \mat{\Lambda}_2 \\
\end{bmatrix}
\ee
 %
 where $\mat{W}_1$ comprises the dominant eigenspace of $\mat{C}$ regarded as
the active subspace, and $\mat{\Lambda}_1$ is the corresponding set of
eigenvalues. Once the active subspace is computed, dimension reduction is
accomplished by transforming the set of canonical random variables, $\bm{\xi}$
into a reduced set, $\mat{W}_1^\top \bm{\xi}$, referred to as the set of active
variables. The number of active variables is thus equal to the number
of dominant eigenvectors. The model output in the original space, $f(\bm{\xi})$,
is expressed in terms of the active variables as $G(\mat{W}_1^\top \bm{\xi})$.
Hence, $f$ is essentially evaluated at the projection of $\bm{\xi}$ on the
column space of $\mat{W}_1$. We could thus confine UQ to the inputs in the
active subspace whose dimension is typically much smaller (in applications that
admit such a subspace) than the dimension of the original input space. To 
expedite UQ, one could easily fit a regression-surface to $G$ using the 
following sequence of steps outlined in~\cite{Constantine:2015} (see chapter 4).
\begin{enumerate}
\item Consider a set of input-output pairs, $\{\bm{\xi}_i, f(\bm{\xi}_i)\}$.
\item For each $\bm{\xi}_i$, compute $\mat{W}_1^\top\bm{\xi}_i$. Note that
 $G(\mat{W}_1^\top \bm{\xi})$ $\approx$ $f(\bm{\xi}_i)$.
\item Fit a regression surface, $\tilde{G}(\mat{W}_1^\top\bm{\xi}_i)$ $\approx$ 
$G(\mat{W}_1^\top \bm{\xi})$.
\item Overall approximation, $f(\bm{\xi})$ $\approx$ $\tilde{G}(\mat{W}_1^\top\bm{\xi})$.
\end{enumerate}

In practice, the matrix, $\mat{C}$ is typically
approximated using pseudo-random sampling techniques such as Monte Carlo and
Latin hypercube sampling. The integral in~\eqref{eq:C} is replaced with a
summation as follows:
 %
 \be
 \mat{C}\approx \hat{\mat{C}} = \frac{1}{N}\sum\limits_{i=1}^{N} 
 (\nabla_{\bm{\xi}}f(\bm{\xi}_i))(\nabla_{\bm{\xi}}f(\bm{\xi}_i))^\top
 = \hat{\mat{W}}\hat{\mat{\Lambda}}\hat{\mat{W}}^\top
\label{eq:chat}
 \ee
 %
Clearly, the amount of computational effort associated with constructing the matrix,
$\hat{\mat{C}}$ scales with the number of samples, $N$. Hence, an iterative
computational approach is adopted in this work to ensure that $N$ is sufficiently
large so that the dominant eigenvectors in $\hat{\mat{W}}_1$ are approximated
with reasonable accuracy as discussed later in Section~\ref{sec:method}.

@@

Combining Proposition~\ref{prp:dgsm_bound} and~\eqref{equ:sobol_bound}, shows
an interesting link between the activity scores and total Sobol' indices.
Specifically, by computing the activity scores, we can identify unimportant
parameters.  
One can attempt to reduce parameter dimension by fixing
unimportant parameters at nominal values. Suppose $\xi_i$ is
deemed unimportant as a result of a global sensitivity analysis
based on computing the activity scores. We want to understand
what is the approximation error that occurs once we fix $\xi_i$ 
at a nominal value.
%Let $\I \subset \{1, \ldots,
%Np\}$ be an index set that indexes the unimportant parameters in the parameter
%vector $\vec{\xi}$.  Letting $|\I|$ denote the number of elements of $\I$, 
%we denote by $\vec{z} \in \R^{|\I|}$ a generic vector of nominal values for
%the unimportant parameters $\vec\xi_\I$, and let $\vec{y}$ be a generic vector of
%important variables, $\vec\xi_{\I^c}$,
%where $\I^c$ denotes the complement of $\I$ in $\{1, \ldots, \Np\}$. 
%For a vector $\xi$, we let $\xin$ denote the vector obtained by removing
%$\xi_i$.

### 2018-08-06 Paper/method.tex
.
..

In this section, we present the computational strategy used for computing the
active subspace in the case of H$_2$/O$_2$ reaction kinetics with uncertain
pre-exponent of individual reaction rates provided in Table~\ref{tab:kinetics}.
Specifically, we employ two strategies: 
A gradient-based strategy that involves computation of model
gradients in order to construct the matrix $\hat{\mat{C}}$
in~\eqref{eq:chat}, and a gradient-free approach 
that involves a (\alennote{should we say 
regression-based?}) local linear approximation of
the model output to circumvent the computational cost associated with gradient
computation. The gradient-based approach is based on Algorithm
1.1, and the gradient-free approach is based on Algorithm 1.2
in~\cite{Constantine:2015}. In this work, we implement the two
strategies in an iterative manner and demonstrate potential for
additional savings in both cases.  For the present application,
our findings based on the two strategies are observed to be consistent.  
Since,
the gradient-free approach is more efficient, this observation
encourages its use for high-dimensional applications as discussed later in
Section~\ref{sec:app}. 
%Note however, that the model output is required to be
%differentiable throughout the input domain in both cases.

@@

As discussed earlier, in the gradient-based approach, the elements of
$\hat{\mat{C}}$ are estimated using model gradients.  In situations where the
gradient is not available analytically, one could consider using finite
difference and other relatively more efficient techniques such as automatic
differentiation~\cite{Kiparissides:2009} and adjoint-based gradient
computation~\cite{Jameson:1988,Gunzburger:2003,Borzi:2011,Alexanderian:2017}.
Model evaluations at neighbouring points are required if finite difference is
used. Hence, for $N$ samples in a $d$-dimensional parameter space, $N(d+1)$
model evaluations are needed.

We begin by evaluating the gradient of the model output, $\nabla_{\bm{\xi}}f$,
at an initial set of $n_1$ samples.  For this purpose, $\bm{\xi}_i$'s are
projected to the physical space as $\bm{\theta}_i$'s. 
\alennote{The mention of $\bm{\theta}$ is redundant. 
Say what $\bm{\xi}_i$, $i = 1, \ldots, n_1$ are before using them.} 
Using the gradient
evaluations, the matrix, $\hat{\mat{C}}$ is computed. Eigenvalue decomposition
of $\hat{\mat{C}}$ yields an initial estimate of the dominant eigenspace,
$\hat{\mat{W}}_1$ and corresponding set of eigenvalues,
$\hat{\mat{\Lambda}}_1$.  At each subsequent iteration, model evaluations are
generated at a new set of $n_s$ samples. The new set of gradient evaluations
are augmented with the available set to re-construct $\hat{\mat{C}}$ followed
by its eigenvalue decomposition. The relative change in the norm of the
difference in squared value of individual components of the dominant
eigenvectors between subsequent iterations is evaluated. The process is
terminated and the resulting eigenspace is considered to be converged once
highest relative change at iteration $k$, $\delta \hat{\mat{W}}_{1,j}^{(k)}$,
is smaller than a given tolerance, $\tau$.  A regression fit to
$G(\hat{\mat{W}}_1^\top\bm{\xi})$ is used as a surrogate to characterize and
quantify the uncertainty in the model output. Moreover, the components of the
dominant eigenvectors are used to compute the activity scores, $\bm{\nu}_r(f)$,
which provide insight into the relative importance of the uncertain inputs.
Note that the index, $r$, corresponds to the number of dominant eigenvectors in
$\hat{\mat{W}}_1$. The sequence of steps for the gradient-based approach are
provided in Algorithm~\ref{alg:grad}.

@@

To motivate our iterative strategy, we implement
Algorithm~\ref{alg:grad} to the 19-dimensional H$_2$/O$_2$ reaction kinetics
problem with uncertain $A_i$'s. For the purpose of verification,
$\hat{\mat{C}}$ is initially constructed using a large set of samples
($N$~=~1000) in the input domain and the corresponding active subspace is
computed. In Figure~\ref{fig:eig_comp}~(left), we illustrate the comparison of
the resulting normalized eigenvalue spectrum ($\lambda_1, \ldots, 
\lambda_{19}$) with the same using a much 
smaller set of samples, $n$~=~$\{20,40,80,120\}$.
Additionally, in Figure~\ref{fig:eig_comp} (right), we compare relative $L_2$
norm of the difference between the first five eigenvectors of $\hat{\mat{C}}$.
\alennote{This last sentence is unclear.}

@@

\alennote{I don't see the point of this paragraph; part of it that 
explains the gradient-free approach is unclear, and the part 
that seems to motivate its use is something to mention in the introduction.
Just say what the gradient-free approach is about, and say that we 
follow an iterative strategies as before.}
As mentioned earlier, the gradient-free approach does not involve computation
of the gradient of the model output. Instead, a least-squares fit is performed
to the local linear regression-based approximation to the model output. While
the gradient-free approach offers a significant potential for additional
computational savings, its ability to reasonably approximate the active
subspace has not been investigated for a wide variety of
problems~\cite{Constantine:2015}. We are motivated by the
potential of this approach in reducing the computational effort
associated with the active subspace computation. Moreover, this effort could
help advance this approach and further motivate its usage for a wide variety of
high-dimensional applications in the future.

First, we implement the gradient-free approach to the 19-dimensional problem and compare our findings with
those obtained using the gradient-based approach. Later, in
Section~\ref{sec:app}, we compute the active subspace for a 33-dimensional application using the gradient-free
approach. Once again, the strategy is implemented in an iterative manner as shown in~\ref{sub:grad}. 
It is observed that the results from the gradient-free approach are consistent with the gradient-based approach
as discussed further below.

In the gradient-free approach, the model output is first computed at an initial set of $n_1$ samples. An independent
set of $M$ samples is also drawn according to the joint probability distribution of the inputs. For each sample
in $M$, $p$ nearest neighbours in the $n_1$ set are identified. A least-squares fit to model evaluations at
the $p$ samples is performed. The slope vector ($\bm{b}_i$) of the regression-fit is stored separately and the process is
repeated for each sample in $M$. The symmetric and positive semi-definite matrix, $\hat{\mat{C}}$ is hence constructed
using the slope vectors as follows:
%
\be
\hat{\mat{C}} = \frac{1}{M}\sum\limits_{i=1}^{M}\bm{b}_i\bm{b}_i^\top = \hat{\bm{W}}\hat{\bm{\Lambda}}\hat{\bm{W}}^\top
\ee
%
Hence, the gradient-free approach essentially differs in the procedure for constructing the matrix, $\hat{\mat{C}}$.
The active subspace computation involves the same sequence of steps as outlined previously for the gradient-based 
approach in Algorithm~\ref{alg:grad}. The procedure for constructing $\hat{\mat{C}}$ using local linear approximation
of the model output is provided in Algorithm~\ref{alg:free2}.

@@

As mentioned earlier, the model output $f(\bm{\xi})$ predominantly varies in the active subspace. Hence, 
$f(\bm{\xi})$ can be transformed as a function of the active variable as $G(\hat{\mat{W}}_1^\top\bm{\xi})$ in
the subspace. The plot of $G$ versus $\hat{\mat{W}}_1^\top\bm{\xi}$, regarded as the \textit{sufficient summary plot} (SSP) is 
compared for the two approaches in Figure~\ref{fig:comp_ssp}.
%
\begin{figure}[htbp]
 \begin{center}
  \includegraphics[width=0.48\textwidth]{./Figures/comp_ssp}
\caption{An illustrative comparison of the SSPs generated using the gradient-based and the gradient-free approaches.}
\label{fig:comp_ssp}
\end{center}
\end{figure}
%
The two SSPs are observed to be in excellent agreement with each other. Moreover, it is interesting to note
that the variability in ignition
delay based on the considered prior marginals for $A_i$'s is observed to be approximately linear in the
active subspace; consistent with the fact the it is 1-dimensional.
\alennote{The last sentence is incomplete; also clarify that ``it'' refers to 
the active subspace.}

We complement the comparative assessment of the two approaches by estimating
the activity scores for individual uncertain parameters ($\alpha_i$) using the
components of the dominant eigenvector as shown in Algorithm~\ref{alg:grad}.
The activity scores (\alennote{relative activity scores?}) for the 19 uncertain pre-exponents ($A_i$'s), estimated
using the gradient-based and gradient-free approaches are plotted in
Figure~\ref{fig:comp_as} (left).  Additionally, for the purpose of
verification, we compare these estimates with those obtained using the
corresponding DGSMs, as reported in~\cite{Vohra:2018}, in
Figure~\ref{fig:comp_as} (right).

%
\begin{figure}[htbp]
 \begin{center}
  \includegraphics[width=0.45\textwidth]{./Figures/comp_as}
  \includegraphics[width=0.45\textwidth]{./Figures/ub_conv_kinetics_rich}
\caption{Left: A bar-graph of activity scores ($\alpha_i$'s) for the 19 uncertain pre-exponents ($A_i$'s).
Right: A plot of the screening metric ($\widehat{\mathcal{C}_i\mu_i}$), defined as a normalized product of the
Poincare constant and the DGSM ($\mu_i$) (adapted from~\cite{Vohra:2018}).}
\label{fig:comp_as}
\end{center}
\end{figure}
%
The activity scores estimated using the two approaches agree favourably with each other as well as those
based on the screening metric involving the DGSMs from~\cite{Vohra:2018}. It is observed that the uncertainty
associated with the ignition delay is largely due to the uncertainty in $A_9$ and $A_1$. Sensitivity towards
$A_{15}$ and $A_{17}$ is also found to be significant.

>>>>

In this section, we present the computational strategy used for computing the
active subspace in the case of H$_2$/O$_2$ reaction kinetics with uncertain
pre-exponent of individual reaction rates provided in Table~\ref{tab:kinetics}.
Specifically, we employ two strategies: 
A gradient-based strategy that involves computation of model
gradients in order to construct the matrix $\hat{\mat{C}}$
in~\eqref{eq:chat}, and a gradient-free approach 
that involves a regression-based local linear approximation of
the model output to circumvent the computational cost associated with gradient
computation. 
The gradient-based approach is based on Algorithm
1.1, and the gradient-free approach is based on Algorithm 1.2
in~\cite{Constantine:2015}. In this work, we implement the two
strategies in an iterative manner and demonstrate potential for
additional savings in both cases.  For the present application,
our findings based on the two strategies are observed to be consistent.  
Since,
the gradient-free approach is more efficient, this observation
encourages its use for high-dimensional applications as discussed later in
Section~\ref{sec:app}. 
%Note however, that the model output is required to be
%differentiable throughout the input domain in both cases.

@@

As discussed earlier, in the gradient-based approach, the elements of
$\hat{\mat{C}}$ are estimated using model gradients.  In situations where the
gradient is not available analytically, one could consider using finite
difference and other relatively more efficient techniques such as automatic
differentiation~\cite{Kiparissides:2009} and adjoint-based gradient
computation~\cite{Jameson:1988,Gunzburger:2003,Borzi:2011,Alexanderian:2017}.
Model evaluations at neighboring points are required if finite difference is
used. Hence, for $N$ samples in a $d$-dimensional parameter space, $N(d+1)$
model evaluations are needed.

We begin by evaluating the gradient of the model output, $\nabla_{\bm{\xi}}f$,
at an initial set of $n_0$ samples denoted by $\bm{\xi}_i$, $i$ = $1,\ldots,n_0$.
Using the gradient
evaluations, the matrix, $\hat{\mat{C}}$ is computed. Eigenvalue decomposition
of $\hat{\mat{C}}$ yields an initial estimate of the dominant eigenspace,
$\hat{\mat{W}}_1$ and a diagonal matrix, 
$\hat{\mat{\Lambda}}_1$ that comprises the eigenvalues along the diagonal.
 At each subsequent iteration, model evaluations are
generated at a new set of $n_k$ samples. The new set of gradient evaluations
are augmented with the available set to re-construct $\hat{\mat{C}}$ followed
by its eigenvalue decomposition. The relative change in the norm of the
difference in squared value of individual components of the dominant
eigenvectors between subsequent iterations is evaluated. The process is
terminated and the resulting eigenspace is considered to be converged once
the maximum relative change at iteration $k$, $\delta \hat{\mat{W}}_{1,j}^{(k)}$
($j$ is used as an index for the eigenvectors),
is smaller than a given tolerance, $\tau$.  A regression fit to
$G(\hat{\mat{W}}_1^\top\bm{\xi})$ is used as a surrogate to characterize and
quantify the uncertainty in the model output. Moreover, the components of the
dominant eigenvectors are used to compute the activity scores, $\bm{\nu}_r(f)$,
which provide insight into the relative importance of the uncertain inputs.
Note that the index, $r$, corresponds to the number of dominant eigenvectors in
$\hat{\mat{W}}_1$. The sequence of steps for the gradient-based approach are
provided in Algorithm~\ref{alg:grad}.

@@

To motivate our iterative strategy, we implement
Algorithm~\ref{alg:grad} to the 19-dimensional H$_2$/O$_2$ reaction kinetics
problem with uncertain $A_i$'s. For the purpose of verification,
$\hat{\mat{C}}$ is initially constructed using a large set of samples
($N$~=~1000) in the input domain and the corresponding active subspace is
computed. In Figure~\ref{fig:eig_comp}~(left), we illustrate the comparison of
the resulting normalized eigenvalue spectrum ($\lambda_1, \ldots, 
\lambda_{19}$) with the same using a much 
smaller set of samples, $n$~=~$\{20,40,80,120\}$.
Additionally, we compare relative L-2
norm of the difference between the squared value of corresponding components
of the eigenvectors, obtained using the large set, $N$, and the smaller sets, $n$.
In  Figure~\ref{fig:eig_comp} (right), we plot the relative L-2 error norm for the
first five eigenvectors.

@@

The gradient-free approach exploits a regression-based local linear
approximation to the model output for estimating the gradients, $\bm{g}^i$
in Algorithm~\ref{alg:grad}. Once again, we use an iterative strategy for
computing the active subspace. The utility of this approach is demonstrated
by constructing an efficient response surface for the 19-dimensional 
H$_2$/O$_2$ reaction kinetics problem considered in~\ref{sub:grad}.
Moreover, we compare our findings with
those obtained using the gradient-based approach to assess consistency
between the two approaches. Motivated by our findings, we compute the active 
subspace for a 33-dimensional H$_2$/O$_2$ reaction kinetics application using the gradient-free
approach in Section~\ref{sec:app}.

In the gradient-free approach, the model output is first computed at an initial set of $n_0$ samples. 
An independent set of $M$ samples is also drawn according to the joint probability distribution of the inputs.
For each sample in $M$, $p$ nearest neighbors in the $n_0$ set are identified. A least-squares fit to model
evaluations at the $p$ samples is performed. The slope vector ($\bm{b}_i$) of the regression-fit is stored
separately and the process is repeated for each sample in $M$. The symmetric and positive semi-definite matrix,
$\hat{\mat{C}}$ is constructed using the slope vectors as follows:
%
\be
\hat{\mat{C}} = \frac{1}{M}\sum\limits_{i=1}^{M}\bm{b}_i\bm{b}_i^\top = \hat{\bm{W}}\hat{\bm{\Lambda}}\hat{\bm{W}}^\top
\ee
%
Hence, the gradient-free approach essentially differs in the procedure for constructing the matrix, $\hat{\mat{C}}$.
The active subspace computation involves the same sequence of steps as outlined previously for the gradient-based 
approach in Algorithm~\ref{alg:grad}. The procedure for constructing $\hat{\mat{C}}$ using a regression-based
local linear approximation
of the model output is provided in Algorithm~\ref{alg:free2}.

@@

As mentioned earlier, the model output $f(\bm{\xi})$ predominantly varies in the active subspace. Hence, 
$f(\bm{\xi})$ can be transformed as a function of the active variable as $G(\hat{\mat{W}}_1^\top\bm{\xi})$ in
the subspace. The plot of $G$ versus $\hat{\mat{W}}_1^\top\bm{\xi}$, regarded as the \textit{sufficient summary plot} (SSP) is 
compared for the two approaches in Figure~\ref{fig:comp_ssp}.
%
\begin{figure}[htbp]
 \begin{center}
  \includegraphics[width=0.48\textwidth]{./Figures/comp_ssp}
\caption{An illustrative comparison of the SSPs generated using the gradient-based and the gradient-free approaches.}
\label{fig:comp_ssp}
\end{center}
\end{figure}
%
The two SSPs are observed to be in excellent agreement with each other. Moreover, it is interesting to note
that the variability in ignition
delay based on the considered prior marginals for $A_i$'s is observed to be approximately linear in the
active subspace. This observation indicates the the active subspace is 1-dimensional in this case.

We complement the comparative assessment of the two approaches by estimating
the activity scores for individual uncertain parameters ($\nu_{i,r}$; $r$=1 since the
active subspace is 1-dimensional)
using the components of the dominant eigenvector as shown in Algorithm~\ref{alg:grad}.
The activity scores for the 19 uncertain pre-exponents ($A_i$'s), estimated
using the gradient-based and gradient-free approaches are plotted in
Figure~\ref{fig:comp_as} (left).  Additionally, for the purpose of
verification, we compare these estimates with those obtained using the
corresponding DGSMs, as reported in~\cite{Vohra:2018}, in
Figure~\ref{fig:comp_as} (right).

%
\begin{figure}[htbp]
 \begin{center}
  \includegraphics[width=0.45\textwidth]{./Figures/comp_as}
  \includegraphics[width=0.45\textwidth]{./Figures/ub_conv_kinetics_rich}
\caption{Left: A bar-graph of activity scores ($\nu_{i,r}$'s) for the 19 uncertain pre-exponents ($A_i$'s); $r$
denotes the number of eigenvectors in the active subspace.
Right: A plot of the screening metric ($\widehat{\mathcal{C}_i\mu_i}$), defined as a normalized product of the
Poincare constant and the DGSM ($\mu_i$) (adapted from~\cite{Vohra:2018}).}
\label{fig:comp_as}
\end{center}
\end{figure}
%
The activity scores estimated using the two approaches agree favorably with each other as well as those
based on the screening metric involving the DGSMs from~\cite{Vohra:2018}. It is observed that the uncertainty
associated with the ignition delay is largely due to the uncertainty in $A_9$ and $A_1$. Sensitivity towards
$A_{15}$ and $A_{17}$ is also found to be significant.
